---
title: "LIME_World_Happiness_Data_Visualization"
author: "Dhaval Mandalia"
date: "April 22, 2019"
output: html_document
---
# Explaining complex machine learning models with LIME
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(OneR)
library(dplyr)
library(tidyverse)

```

I am using the [World Happiness Report](https://www.kaggle.com/unsdsn/world-happiness) from Kaggle.

```{r 1.Load Data}
data_17 <- read.table("data/world-happiness-2017.csv", sep = ",", header = TRUE)
data_16 <- read.table("data/world-happiness-2016.csv", sep = ",", header = TRUE)
data_15 <- read.table("data/world-happiness-2015.csv", sep = ",", header = TRUE)
```

In the 2016 data there are upper and lower CI for the happiness score given, while in the 2015 data we have standard errors. Because I want to combine data from the two years, I am using only columns that are in both datasets.

```{r 2.Combine Data}
common_feats <- colnames(data_16)[which(colnames(data_16) %in% colnames(data_15))]

# features and response variable for modeling
feats <- setdiff(common_feats, c("Country", "Happiness.Rank", "Happiness.Score"))
response <- "Happiness.Score"

# combine data from 2015 and 2016
data_15_16 <- rbind(select(data_15, one_of(c(feats, response))),
              select(data_16, one_of(c(feats, response))))

```
The response variable happiness score is on a numeric scale. OneR could also perform regression but here, I want to compare classification tasks. For classifying happiness, I create three bins for low, medium and high values of the happiness score. In order to not having to deal with unbalanced data, I am using the bin() function from OneR with method = "content". For plotting the cut-points, I am extracting the numbers from the default level names.

```{r 3.Process Data}
data_15_16$Happiness.Score.l <- bin(data_15_16$Happiness.Score, nbins = 3, method = "content")

intervals <- paste(levels(data_15_16$Happiness.Score.l), collapse = " ")
intervals <- gsub("\\(|]", "", intervals)
intervals <- gsub(",", " ", intervals)
intervals <- as.numeric(unique(strsplit(intervals, " ")[[1]]))

data_15_16 %>%
  ggplot() +
    geom_density(aes(x = Happiness.Score), color = "blue", fill = "blue", alpha = 0.4) +
    geom_vline(xintercept = intervals[2]) +
    geom_vline(xintercept = intervals[3])
```
Now removing the original happiness score column from the data for modeling and rename the factor levels of the response variable.
Because there are only 9 features in this small dataset, I want to explore them all individually before modeling. First, I am plotting the only categorical variable: Region.

This plots shows that there are a few regions with very strong biases in happiness: People in Western Europe, Australia, New Zealand, North America, Latin American and the Caribbean tend to me in the high happiness group, while people in sub-saharan Africa and Southern Asia tend to be the least happiest.
```{r 4.Visualize Data 1}
data_15_16 <- select(data_15_16, -Happiness.Score) %>%
  mutate(Happiness.Score.l = plyr::revalue(Happiness.Score.l, c("(2.83,4.79]" = "low", "(4.79,5.89]" = "medium", "(5.89,7.59]" = "high")))
data_15_16 %>%
  ggplot(aes(x = Region, fill = Happiness.Score.l)) +
    geom_bar(position = "dodge", alpha = 0.7) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          plot.margin = unit(c(0, 0, 0, 1.5), "cm")) +
    scale_fill_brewer(palette = "Set1")
```
The remaining quantitative variables show happiness biases to varying degrees: e.g. low health and life expectancy is strongly biased towards low happiness, economic factors, family and freedom show a bias in the same direction, albeit not as strong.
While OneR could also handle categorical data, in this example, I only want to consider the quantitative features to show the differences between OneR and other machine learning algorithms.
```{r 5.Visualize Data 2}
data_15_16 %>%
  gather(x, y, Economy..GDP.per.Capita.:Dystopia.Residual) %>%
  ggplot(aes(x = y, fill = Happiness.Score.l)) +
    geom_histogram(alpha = 0.7) +
    facet_wrap(~ x, scales = "free", ncol = 4) +
    scale_fill_brewer(palette = "Set1")

data_15_16 <- select(data_15_16, -Region)
```

## Modelling
The algorithms I will compare to OneR will be run via the caret package. Also use caret’s createDataPartition() function to partition the data into training (70%) and test sets (30%).
OneR only accepts categorical features. Because we have numerical features, we need to convert them to factors by splitting them into appropriate bins. While the original OneR algorithm splits the values into ever smaller factors, this has been changed in this R-implementation with the argument of preventing overfitting. We can either split the data into pre-defined numbers of buckets (by length, content or cluster) or we can use the optbin() function to obtain the optimal number of factors from pairwise logistic regression or information gain.
```{r 6.Modelling Data 1}
set.seed(42)
index <- createDataPartition(data_15_16$Happiness.Score.l, p = 0.7, list = FALSE)
train_data <- data_15_16[index, ]
test_data  <- data_15_16[-index, ]

# default method length
data_1 <- bin(train_data, nbins = 5, method = "length")

# method content
data_2 <- bin(train_data, nbins = 5, method = "content")

# method cluster
data_3 <- bin(train_data, nbins = 3, method = "cluster")

# optimal bin number logistic regression
data_4 <- optbin(formula = Happiness.Score.l ~., data = train_data, method = "logreg")

# optimal bin number information gain
data_5 <- optbin(formula = Happiness.Score.l ~., data = train_data, method = "infogain")
```


Now running the OneR models. During model building, the chosen attribute/feature with highest accuracy along with the top 7 features decision rules and accuracies are printed. Unfortunately, this information is not saved in the model object; this would have been nice in order to compare the importance of features across models later on.

Here, all five models achieved highest prediction accuracy with the feature Economy GDP per capita.
```{r 7.Modelling Data 2}
for (i in 1:5) {
  data <- get(paste0("data_", i))
  print(model <- OneR(formula = Happiness.Score.l ~., data = data, verbose = TRUE))
  assign(paste0("model_", i), model)
}
## 
##     Attribute                     Accuracy
## 1 * Economy..GDP.per.Capita.      63.96%  
## 2   Health..Life.Expectancy.      59.91%  
## 3   Family                        57.21%  
## 4   Dystopia.Residual             51.8%   
## 5   Freedom                       49.55%  
## 6   Trust..Government.Corruption. 45.5%   
## 7   Generosity                    41.89%  
## ---
## Chosen attribute due to accuracy
## and ties method (if applicable): '*'
## 
## 
## Call:
##OneR(data_1, data = data, formula = Happiness.Score.l ~ ., verbose = TRUE)
## 
## Rules:
## If Economy..GDP.per.Capita. = (-0.00182,0.365] then Happiness.Score.l = low
## If Economy..GDP.per.Capita. = (0.365,0.73]     then Happiness.Score.l = low
## If Economy..GDP.per.Capita. = (0.73,1.09]      then Happiness.Score.l = medium
## If Economy..GDP.per.Capita. = (1.09,1.46]      then Happiness.Score.l = high
## If Economy..GDP.per.Capita. = (1.46,1.83]      then Happiness.Score.l = high
## 
## Accuracy:
## 142 of 222 instances classified correctly (63.96%)
## 
## 
##     Attribute                     Accuracy
## 1 * Economy..GDP.per.Capita.      64.41%  
## 2   Health..Life.Expectancy.      60.81%  
## 3   Family                        59.91%  
## 4   Trust..Government.Corruption. 55.41%  
## 5   Freedom                       53.15%  
## 5   Dystopia.Residual             53.15%  
## 7   Generosity                    41.44%  
## ---
## Chosen attribute due to accuracy
## and ties method (if applicable): '*'
## 
## 
## Call:
## OneR(data_2, data = data, formula = Happiness.Score.l ~ ., verbose = TRUE)
## 
## Rules:
## If Economy..GDP.per.Capita. = (-0.00182,0.548] then Happiness.Score.l = low
## If Economy..GDP.per.Capita. = (0.548,0.877]    then Happiness.Score.l = low
## If Economy..GDP.per.Capita. = (0.877,1.06]     then Happiness.Score.l = medium
## If Economy..GDP.per.Capita. = (1.06,1.28]      then Happiness.Score.l = medium
## If Economy..GDP.per.Capita. = (1.28,1.83]      then Happiness.Score.l = high
## 
## Accuracy:
## 143 of 222 instances classified correctly (64.41%)
## 
## 
##     Attribute                     Accuracy
## 1 * Economy..GDP.per.Capita.      63.51%  
## 2   Health..Life.Expectancy.      62.16%  
## 3   Family                        54.5%   
## 4   Freedom                       50.45%  
## 4   Dystopia.Residual             50.45%  
## 6   Trust..Government.Corruption. 43.24%  
## 7   Generosity                    36.49%  
## ---
## Chosen attribute due to accuracy
## and ties method (if applicable): '*'
## 
## 
## Call:
## OneR(data_3, data = data, formula = Happiness.Score.l ~ ., verbose = TRUE)
## 
## Rules:
## If Economy..GDP.per.Capita. = (-0.00182,0.602] then Happiness.Score.l = low
## If Economy..GDP.per.Capita. = (0.602,1.1]      then Happiness.Score.l = medium
## If Economy..GDP.per.Capita. = (1.1,1.83]       then Happiness.Score.l = high
## 
## Accuracy:
## 141 of 222 instances classified correctly (63.51%)
## 
## 
##     Attribute                     Accuracy
## 1 * Economy..GDP.per.Capita.      63.96%  
## 2   Health..Life.Expectancy.      62.16%  
## 3   Family                        58.56%  
## 4   Freedom                       51.35%  
## 5   Dystopia.Residual             50.9%   
## 6   Trust..Government.Corruption. 46.4%   
## 7   Generosity                    40.09%  
## ---
## Chosen attribute due to accuracy
## and ties method (if applicable): '*'
## 
## 
## Call:
## OneR(data_4, data = data, formula = Happiness.Score.l ~ ., verbose = TRUE)
## 
## Rules:
## If Economy..GDP.per.Capita. = (-0.00182,0.754] then Happiness.Score.l = low
## If Economy..GDP.per.Capita. = (0.754,1.12]     then Happiness.Score.l = medium
## If Economy..GDP.per.Capita. = (1.12,1.83]      then Happiness.Score.l = high
## 
## Accuracy:
## 142 of 222 instances classified correctly (63.96%)
## 
## 
##     Attribute                     Accuracy
## 1 * Economy..GDP.per.Capita.      67.12%  
## 2   Health..Life.Expectancy.      65.77%  
## 3   Family                        61.71%  
## 4   Trust..Government.Corruption. 56.31%  
## 5   Dystopia.Residual             55.41%  
## 6   Freedom                       50.9%   
## 7   Generosity                    43.69%  
## ---
## Chosen attribute due to accuracy
## and ties method (if applicable): '*'
## 
## 
## Call:
## OneR(data_5, data = data, formula = Happiness.Score.l ~ ., verbose = TRUE)
## 
## Rules:
## If Economy..GDP.per.Capita. = (-0.00182,0.68] then Happiness.Score.l = low
## If Economy..GDP.per.Capita. = (0.68,1.24]     then Happiness.Score.l = medium
## If Economy..GDP.per.Capita. = (1.24,1.83]     then Happiness.Score.l = high
## 
## Accuracy:
## 149 of 222 instances classified correctly (67.12%)

```

## Model evaluation
The function eval_model() prints confusion matrices for absolute and relative predictions, as well as accuracy, error and error rate reduction. For comparison with other models, it would have been convenient to be able to extract these performance metrics directly from the eval_model object, instead of only the confusion matrix and values of correct/all instances and having to re-calculate performance metrics again manually.
Because I want to calculate performance measures for the different classes separately and like to have a more detailed look at the prediction probabilities I get from the models, I prefer to obtain predictions with type = "prob. While I am not looking at it here, this would also allow me to test different prediction thresholds.
```{r 8.Model Evaluation}
for (i in 1:5) {
  model <- get(paste0("model_", i))
  eval_model(predict(model, test_data), test_data$Happiness.Score.l)
}


for (i in 1:5) {
  model <- get(paste0("model_", i))
  pred <- data.frame(model = paste0("model_", i),
                     sample_id = 1:nrow(test_data),
                     predict(model, test_data, type = "prob"),
                     actual = test_data$Happiness.Score.l)
  pred$prediction <- colnames(pred)[3:5][apply(pred[, 3:5], 1, which.max)]
  pred$correct <- ifelse(pred$actual == pred$prediction, "correct", "wrong")
  pred$pred_prob <- NA
  
  for (j in 1:nrow(pred)) {
    pred[j, "pred_prob"] <- max(pred[j, 3:5])
  }
  
  if (i == 1) {
    pred_df <- pred
  } else {
    pred_df <- rbind(pred_df, pred)
  }
}
```

## Comparing other algorithms

### Decision trees
Economy GDP per capita is the second highest node here, the best predictor here would be health and life expectancy.

```{r 9.Decision Trees 1}
library(rpart)
library(rpart.plot)

set.seed(42)
fit <- rpart(Happiness.Score.l ~ .,
            data = train_data,
            method = "class",
            control = rpart.control(xval = 10), 
            parms = list(split = "information"))

rpart.plot(fit, extra = 100)
```
In order to compare the models, I am producing the same output table for predictions from this model and combine it with the table from the OneR models.

```{r 9.Decision Trees 2}
pred <- data.frame(model = "rpart",
                   sample_id = 1:nrow(test_data),
                   predict(fit, test_data, type = "prob"),
                   actual = test_data$Happiness.Score.l)
  pred$prediction <- colnames(pred)[3:5][apply(pred[, 3:5], 1, which.max)]
  pred$correct <- ifelse(pred$actual == pred$prediction, "correct", "wrong")
  pred$pred_prob <- NA
  
  for (j in 1:nrow(pred)) {
    pred[j, "pred_prob"] <- max(pred[j, 3:5])
  }
  
pred_df_final <- rbind(pred_df,
                       pred)
```

### Random Forest
Training a Random Forest model.The varImp() function from caret shows us which feature was of highest importance for the model and its predictions.

Here, we again find Economy GDP per captia on top.
```{r 11.Random Forest 1}
set.seed(42)
model_rf <- caret::train(Happiness.Score.l ~ .,
                         data = train_data,
                         method = "rf",
                         trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 5, 
                                                  verboseIter = FALSE))
varImp(model_rf)
pred <- data.frame(model = "rf",
                   sample_id = 1:nrow(test_data),
                   predict(model_rf, test_data, type = "prob"),
                   actual = test_data$Happiness.Score.l)
  pred$prediction <- colnames(pred)[3:5][apply(pred[, 3:5], 1, which.max)]
  pred$correct <- ifelse(pred$actual == pred$prediction, "correct", "wrong")
  pred$pred_prob <- NA
  
  for (j in 1:nrow(pred)) {
    pred[j, "pred_prob"] <- max(pred[j, 3:5])
  }
pred_df_final <- rbind(pred_df_final,
                       pred)
```

### Extreme gradient boosting trees
Gradient boosting is another decision tree-based algorithm. As before, we find Dystopia.Residual as most important feature.

```{r 12.XGB 1}
set.seed(42)
model_xgb <- caret::train(Happiness.Score.l ~ .,
                         data = train_data,
                         method = "xgbTree",
                         trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 5, 
                                                  verboseIter = FALSE))
varImp(model_xgb)

```
```{r 13.XGB 2}
pred <- data.frame(model = "xgb",
                   sample_id = 1:nrow(test_data),
                   predict(model_xgb, test_data, type = "prob"),
                   actual = test_data$Happiness.Score.l)
  pred$prediction <- colnames(pred)[3:5][apply(pred[, 3:5], 1, which.max)]
  pred$correct <- ifelse(pred$actual == pred$prediction, "correct", "wrong")
  pred$pred_prob <- NA
  
  for (j in 1:nrow(pred)) {
    pred[j, "pred_prob"] <- max(pred[j, 3:5])
  }

pred_df_final <- rbind(pred_df_final,
                       pred)
```
### Neural network

```{r 14.Neural Network 1}
set.seed(42)
model_nn <- caret::train(Happiness.Score.l ~ .,
                         data = train_data,
                         method = "mlp",
                         trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 5, 
                                                  verboseIter = FALSE))
varImp(model_nn)

```

```{r 15.Neural Network 2}
pred <- data.frame(model = "nn",
                   sample_id = 1:nrow(test_data),
                   predict(model_nn, test_data, type = "prob"),
                   actual = test_data$Happiness.Score.l)
  pred$prediction <- colnames(pred)[3:5][apply(pred[, 3:5], 1, which.max)]
  pred$correct <- ifelse(pred$actual == pred$prediction, "correct", "wrong")
  pred$pred_prob <- NA
  
  for (j in 1:nrow(pred)) {
    pred[j, "pred_prob"] <- max(pred[j, 3:5])
  }

pred_df_final <- rbind(pred_df_final,
                       pred)
```


### Model comparisons
The first plot below shows the prediction probabilites for the three happiness levels low, medium and high for each test data instance. For each instance, only the prediction probability of the predicted class (i.e. with the highest value) is shown. The upper row shows correct predictions, the lower row shows wrong predictions.

Sometimes, it is obvious from such a plot if a more stringent prediction threshold could improve things (when wrong predictions tend to be close to the threshold). With three classes to predict, this is obviously not as trivial as if we only had two but the same principle holds true: the smaller the prediction probability, the more uncertain it tends to be.

```{r 16.Model Comparison 1}
pred_df_final %>%
  ggplot(aes(x = actual, y = pred_prob, fill = prediction, color = prediction)) +
    geom_boxplot(alpha = 0.7) +
    facet_grid(correct ~ model) +
    scale_color_brewer(palette = "Set1") +
    scale_fill_brewer(palette = "Set1")
```
Probably the most straight-forwards performance measure is accuracy: i.e. the proportion of correct predictions vs the total number of instances to predict. The closer to 1, the better the accuracy.

Not surprisingly, the more complex models tend to be more accurate - albeit only slightly.
```{r 17.Model Comparison 2}
pred_df_final %>%
  group_by(model) %>%
  dplyr::summarise(correct = sum(correct == "correct")) %>%
  mutate(accuracy = correct / nrow(test_data)) %>%
  ggplot(aes(x = model, y = accuracy, fill = model)) +
    geom_bar(stat = "identity") +
    scale_fill_brewer(palette = "Set1")
```
When we look at the three classes individually, it looks a bit more complicated but most models achieved highest accuracy for class “high”.
```{r 18.Model Comparison 2}
pred_df_final %>%
  group_by(model, prediction) %>%
  dplyr::summarise(correct = sum(correct == "correct"),
            n = n()) %>%
  mutate(accuracy = correct / n) %>%
  ggplot(aes(x = model, y = accuracy, fill = prediction)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_brewer(palette = "Set1")
```
### LIME
Run expalanation modules.

```{r 19.Data Setup}
common_feats <- colnames(data_16)[which(colnames(data_16) %in% colnames(data_15))]
# features and response variable for modeling
feats <- setdiff(common_feats, c("Country", "Happiness.Rank", "Happiness.Score"))
response <- "Happiness.Score"

# combine data from 2015 and 2016

data_15_16 <- rbind(dplyr::select(data_15, one_of(c(feats, response))),
                    dplyr::select(data_16, one_of(c(feats, response))))
library(caret)

set.seed(42)
data_15_16$Happiness.Score.l <- bin(data_15_16$Happiness.Score, nbins = 3, method = "content")
index <- createDataPartition(data_15_16$Happiness.Score.l, p = 0.7, list = FALSE)
train_data <- data_15_16[index, ]
test_data  <- data_15_16[-index, ]
```

Run Neural Network model

```{r 20.Model}
set.seed(42)
model_nn <- caret::train(Happiness.Score.l ~ .,
                         data = train_data,
                         method = "mlp",
                         trControl = trainControl(method = "repeatedcv", 
                                                  number = 10, 
                                                  repeats = 5, 
                                                  verboseIter = FALSE))

pred <- data.frame(model = "nn",
                   sample_id = 1:nrow(test_data),
                   predict(model_nn, test_data, type = "prob"),
                   actual = test_data$Happiness.Score.l)
  pred$prediction <- colnames(pred)[3:5][apply(pred[, 3:5], 1, which.max)]
  pred$correct <- ifelse(pred$actual == pred$prediction, "correct", "wrong")
  pred$pred_prob <- NA
  
  for (j in 1:nrow(pred)) {
    pred[j, "pred_prob"] <- max(pred[j, 3:5])
  }




```

Create LIME Explainer

```{r 21.LIME Explainer 1}
library(lime)
library(tidyverse)
explainer <- lime(train_data, model_nn, bin_continuous = TRUE, n_bins = 5, n_permutations = 1000)


pred_cor <- filter(pred, correct == "correct")
pred_wrong <- filter(pred, correct == "wrong")

test_data_cor <- test_data %>%
  mutate(sample_id = 1:nrow(test_data)) %>%
  filter(sample_id %in% pred_cor$sample_id) %>%
  sample_n(size = 3) %>%
  remove_rownames() %>%
  tibble::column_to_rownames(var = "sample_id") %>%
  select(-Happiness.Score.l)

test_data_wrong <- test_data %>%
  mutate(sample_id = 1:nrow(test_data)) %>%
  filter(sample_id %in% pred_wrong$sample_id) %>%
  sample_n(size = 3) %>%
  remove_rownames() %>%
  tibble::column_to_rownames(var = "sample_id") %>%
  select(-Happiness.Score.l)



explanation_cor <- explain(test_data_cor, explainer, n_labels = 3, n_features = 5)
explanation_wrong <- explain(test_data_wrong, explainer, n_labels = 3, n_features = 5)
```







